---
title: "Lab 7 Preprocessing Maps"
output: pdf_document
date: "2026-01-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Maps

By now, you should feel comfortable creating basic interactive charts in python. This week, we’ll step up the complexity and work with maps. Maps can be quite time consuming to produce.

Before you start plotting, think about your data:

What kind of location data do I have? (e.g., latitude/longitude, postcodes, addresses, shapefiles, GeoJSON)

What format does my chosen plotting package require?

Would a different package make this easier? (and what format does that one need?)

## Imports

```{r eval = TRUE, echo = TRUE}

library(dplyr)
library(plotly)
library(ggplot2)
library(googlePolylines) #Introduced this week

file_id <- "1ymbNqfv9s6YGZzN93HFKAhjg0Z5xZXV1"
url <- paste0("https://drive.google.com/uc?id=", file_id)

df <- read.csv(url)

```

## 1. Find data

Martins runs only data has a column called maps.

- What are the types of data contained? What other observations can you make? Consider how to decode this information. Print a single run. Notice the number of escapes `\` appearing.

```{r answer, echo = FALSE, include = FALSE, eval = FALSE}
# JSON like data with single quotes and backslashes including a polyline
# Just a data point for now
fld <- df$map[1]
# split on single quotes
parts <- strsplit(fld, "'")[[1]]
print(parts[8])
```

## 2. Create dataframe

Taking a single datapoint from Martins Runs can you create a dataframe of longditudes and latitudes.

>**HINT**:
>
> You will probably use google Polylines.
>
> Break into steps based on your answer to 1).
>
> The polyline may now include characters such as '\\' gsub is a great method for cleaning ie `gsub("\\\\", "\\", poly, fixed = TRUE)`

```{r coords, echo = FALSE, eval = TRUE}
fld <- df$map[1]
# split on single quotes
parts <- strsplit(fld, "'")[[1]]
#print(parts) Useful print
# 4th piece is the polyline
poly <- parts[8]
# The string in the previous naturally escapes the backslashes, we need to compensate. You can check this with;
# print(poly)
poly <- gsub("\\\\", "\\", poly, fixed = TRUE)

# Decode
coords <- googlePolylines::decode(poly)[[1]]  # data.frame with columns 'lon' and 'lat'

head(coords,5)
```

## 3. Last 20 Runs (Optional)

You can just work with the above dataframe if you prefer. The following is a challenge! If you do attempt it, ensure you understand what is going on.

Retrieve the most recent 20 runs and store them in a DataFrame similar to the previous one, but with an additional column for date. Each point should have a date, so the date will appear multiple times.

```{r twentyruns, echo = FALSE, eval = TRUE}
# Most recent 20 runs
last20 <- df %>%
  mutate(start_date = as.POSIXct(start_date, "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")) %>%
  arrange(desc(start_date)) %>%
  slice_head(n = 20)

# Date labels
date_lab <- format(last20$start_date, "%d-%m-%Y")

coords_df <- bind_rows(lapply(seq_len(nrow(last20)), function(i) {
  fld <- last20$map[i]
  parts <- strsplit(fld, "'")[[1]]
  poly <- parts[8]
  poly <- gsub("\\\\", "\\", poly, fixed = TRUE)

  coords <- googlePolylines::decode(enc2utf8(poly))[[1]]
  coords$date <- date_lab[i]
  coords
}))

head(coords_df)
```

## 4. Save as csv

```{r eval = FALSE, echo = TRUE}
write.csv(coords_df, "map.csv", row.names = FALSE)
```