---
title: "Lab 1 Preprocessing"
output: pdf_document
date: "2026-01-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 1 Data Preprocessing

In this week's exercise, we will check our data and prepare it for visualisation.

## Goals

- Calculate total distance per month.
- Save into a new csv.

## 1. Importing data

We can import Martins csv, or use the follwing code to download the same dataset.

```{r echo = TRUE, eval = TRUE}
file_id <- "1ymbNqfv9s6YGZzN93HFKAhjg0Z5xZXV1"
url <- paste0("https://drive.google.com/uc?id=", file_id)

df <- read.csv(url)
```

## 2. Saving to CSV

If we downloaded the data we can re-save to csv with;

```{r echo = TRUE, eval = FALSE}
write.csv(df, "runningdata.csv", row.names = FALSE)
```

If you are unsure where the file is saved to you can check the working directory with;

```{r echo = TRUE, eval = FALSE}
getwd()
```

You can open and investigate the csv in excel, text editor or spreadsheet program of your choice.

## 3. Explore the dataframe in the viewer

Double click `df` in the environment pane usually on the right to open the data in the viewer

[!bardata](Images/barchartdata.png)

## 4. Get the column headers programatically.

```{r eval=TRUE, echo = FALSE}
colnames(df)
```
## 5. Create a new column containing only the year and month from the date data.

You have choices in how to do this, I recommend you avoid packages at this point and use `as.POSIXct`.
https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/as.POSIX*

You can follow the previous step with `format`.
https://www.geeksforgeeks.org/r-language/how-to-use-date-formats-in-r/

```{r eval = TRUE, echo = FALSE}

# Convert start_date to datetime
df$Date <- as.POSIXct(df$start_date, tz = "UTC")
df$YearMonth <- format(df$Date, "%Y-%m")
head(df$YearMonth)
```
## 6. Aggregate data

Using `aggregate(response ~ group, data = df, FUN = ...)`, create a new dataframe that groups runs by YearMonth and sums Distance to calculate the total distance run each month.

```{r eval = TRUE, echo = FALSE}
monthly <- aggregate(distance ~ YearMonth, data = df, sum)
head(monthly)
```

## 7. Save data to csv

```{r eval = FALSE, echo = TRUE}
write.csv(monthly, "monthly_stats.csv", row.names = FALSE)
```