---
title: "Lab 2 Preprocessing"
output: pdf_document
date: "2026-01-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 2 Preprocessing

## Goal

Create a Scatter Plot to analyze the relationship between Distance and Pace.

## Importing data

We can import Martins csv, or use the following code to download the same dataset.

```{r echo = TRUE, eval = TRUE}
file_id <- "1ymbNqfv9s6YGZzN93HFKAhjg0Z5xZXV1"
url <- paste0("https://drive.google.com/uc?id=", file_id)

df <- read.csv(url)
```

## 1.Create Distance and Pace Columns

Create a new DataFrame that contains:

- Distance in kilometres (km)

- Pace in minutes per kilometre (min/km)

> **Reminder:**  
> Convert units before calculating pace.

```{r eval=TRUE, echo = FALSE}
distancevpace <- data.frame(
  Distance_km = df$distance / 1000, #convert to km
  Pace_min_km = (1000 / df$average_speed) / 60 #convert to km/minute.
)

head(distancevpace)
```

## Save csv

Save the dataframe to csv

```{r eval = FALSE, echo = TRUE}
write.csv(distancevpace, "distancevpace.csv", row.names = FALSE)
```