---
title: "Lab 4 Pie Chart Preprocessing and conditional logic"
output: pdf_document
date: "2026-01-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 4 Preprocessing

## Goal

Create a Pie Chart to show the distribution of runs by Time of Day.

## Imports

```{r eval = TRUE, echo = TRUE}
library(dplyr) # We will now use dplyr as standard

file_id <- "1ymbNqfv9s6YGZzN93HFKAhjg0Z5xZXV1"
url <- paste0("https://drive.google.com/uc?id=", file_id)

df <- read.csv(url)

```

## Create frequency table for runs

We want to create a dataframe that counts how many days contain runs in the AM and PM. Since we cannot assume there is only one run per day, we also include a third category called Both, for days where runs occurred in both the morning and afternoon/evening.

Using either the steps below or your own method, create a summary table showing the number of days in each category (AM, PM, and Both).

### 1. Create new columns for `datetime`, `Date`, `AMPM`

We first parse the `start_date` string into a proper date–time column called `datetime` using:

`as.POSIXct(start_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")`

Once we have `datetime`, we can:

- Use `as.Date(datetime)` to extract the calendar date (year–month–day). This can then be used to group activities by day.
- Use `format(datetime, "%p")` to extract whether the activity occurred in the AM or PM.

In the model answer below, I use these steps as a template to build the required summary table.


```{r echo = FALSE, eval = FALSE}
daily_summary <- df %>%
  mutate(
    datetime = #Your answer
    Date = #Your answer
    AMPM = #Your answer
  ) %>%
  select(Date, AMPM) #Just to show the relevant fields.
head(daily_summary)
```

```{r echo = FALSE, eval = TRUE}
daily_summary <- df %>%
  mutate(
    datetime = as.POSIXct(start_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    Date = as.Date(datetime),
    AMPM = format(datetime, "%p")
  ) %>%
  select(Date, AMPM)
head(daily_summary)
```

### 2. Group by

We can pipe `%>%` our data into `group_by(Date)` to group activities by day. This tags each row with a group label based on the `Date` column, but it does not perform any calculations by itself. Calculations are typically performed afterwards using functions such as `summarise()`.

```{r echo = FALSE, eval = TRUE}
daily_summary <- df %>%
  mutate(
    datetime = as.POSIXct(start_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    Date = as.Date(datetime),
    AMPM = format(datetime, "%p")
  ) %>%
  group_by(Date) %>%
  select(Date, AMPM)
head(daily_summary)
```

### 3. Conditional Logic

We now want to create a daily summary, but we also need to apply a condition. One approach is to count how many unique values of `AMPM` occur on each day. If there are two (i.e., both AM and PM), we label that day as "Both".

In R, we can apply conditional logic using:

`ifelse(condition, value_if_true, value_if_false)`

We use this inside `summarise()` as `summarise()` collapses multiple rows per day into a single row by calculating summary values (i.e., aggregating the data).

Using the example below (or your own method), apply this logic to your dataframe.


```{r echo = TRUE, eval = TRUE}
# Quick example
example <- data.frame(
  Date = as.Date(c("2025-07-19","2025-07-19","2025-07-20","2025-07-21","2025-07-21")),
  AMPM = c("AM", "PM", "AM", "PM", "PM")
) %>%
  group_by(Date) %>%
  summarise(
    period = ifelse(n_distinct(AMPM) == 2, "Both", first(AMPM)),
    .groups = "drop"
  )

example

```
The actual output you should aim for is below;

```{r echo = FALSE, eval = TRUE}

daily_summary <- df %>%
  mutate(
    datetime = as.POSIXct(start_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    Date = as.Date(datetime),
    AMPM = format(datetime, "%p")  # "AM" or "PM"
  ) %>%
  group_by(Date) %>%
  summarise(
    period = ifelse(n_distinct(AMPM) == 2, "Both", first(AMPM)),
    .groups = "drop"
  )
head(daily_summary)

```

### 4. Provide counts

Provide counts. This can be done in a single operation.

```{r echo = FALSE, eval = TRUE}
daily_summary <- df %>%
  mutate(
    datetime = as.POSIXct(start_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    Date = as.Date(datetime),
    AMPM = format(datetime, "%p")  # "AM" or "PM"
  ) %>%
  group_by(Date) %>%
  summarise(
    period = ifelse(n_distinct(AMPM) == 2, "Both", first(AMPM)),
    .groups = "drop"
  ) %>%
  count(period)

daily_summary
```

### 5. Save to csv

