---
title: "Lab 1 Bar Chart Visualisation"
output: pdf_document
date: "2026-01-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 1 Bar Chart Visualisation

In this week excersise we will create a categorical chart with year and month as the categories and distance as the variable.

## Importing data

We can import the csv created in the preprocessing, or use the following code to download an already made csv.

```{r echo = TRUE, eval = TRUE}
file_id <- "1MLylJ6cizJlTFVi_FdiF11pjpT3brlsO"
url <- paste0("https://drive.google.com/uc?id=", file_id)

monthly_stats <- read.csv(url)
head(monthly_stats)
```

## 1. Plot Distance by Month

Using `barplot`, create a bar chart that shows the total distance run in each month.

```{r echo = FALSE, eval = TRUE}
barplot(monthly_stats$TotalDistance/1000,
        names.arg = monthly_stats$YearMonth,
)
```

## 2. Add Titles

Axis labels should include units unless there is a good reason not to. For crowded month labels on a bar chart, you can rotate the x-axis labels using `las` (from `par()` graphical parameters) and shrink them using `cex.names` to improve readability and prevent the axis title being hidden. The `barplot()` arguments `main`, `xlab`, and `ylab` set the plot title and axis labels.

https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/barplot

https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/par

```{r echo = FALSE, eval = TRUE}
barplot(monthly_stats$TotalDistance/1000,
        names.arg = monthly_stats$YearMonth,
        main = "Total Distance Run Per Month",
        xlab = "Year-Month",
        ylab = "Distance(km)",
        las = 2,
        cex.names = 0.7
)
```